Anomalous waves triggered by abrupt depth changes: laboratory experiments and truncated KdV statistical mechanics

### 2019-10-15 paper2_JNLS/paper.tex
NM

\title{Anomalous waves triggered by abrupt depth changes: laboratory experiments and truncated KdV theory}

@@

A recent series of laboratory \cite{bolles2019} and numerical investigations \cite{viotti2014} have demonstrated the emergence of anomalous wave statistics from abrupt variations in bottom topography. Since topographical variations are strictly one-dimensional, these studies can be viewed as offering a bare minimum set of conditions capable of generating anomalous wave statistics. In particular, the more complex mechanism of focusing by 2D bottom topography absent. The studies thus offer a paradigm system, with emergent anomalous statistical properties similar to that seen in more complex systems, but in a tractable context that is amenable to analysis.

@@

We consider the surface displacement $\eta(x,t)$ of unidirectional, shallow-water waves in a reference frame moving with the characteristic wave speed, $\xi = x - ct$. Here, $c = \sqrt{g \depth}$ is leading-order approximation to the wave speed (i.e.~from linear theory), where $g$ gravity and $\depth$ is the local depth.
The leading-order dynamics are corrected to first order in small amplitude by the Korteweg–de Vries equation (KdV), which in dimensional form is given by \cite{whitham2011linear}
\begin{equation}
\label{KdV}
\eta_t + \frac{3 c}{2 \depth} \eta \eta_{\xi} + \frac{c \depth^2}{6} \eta_{\xi \xi \xi} = 0
\end{equation}
% Given by Whitham on p. 461, Eq. 13.98
Motivated by the experiments, we consider waves that originate from a region of constant depth, encounter an abrupt depth change, and continue into another region of constant depth. Thus, depth will be piecewise constant
\begin{align}
\depth = 
\begin{cases}
\dup \quad \mbox{if } x<0 \\
\ddn \quad \mbox{if } x>0
\end{cases}
\end{align}
Most often, we will consider waves moving into shallower depth, so that $\dup > \ddn$.

@@

With the above choices, the dimensionless KdV takes the form
\begin{align}
\label{dimlessKdV}
&u_t + C_3 \drat^{-3/2} \, u u_x + C_2 \drat^{1/2} \, u_{xxx} = 0
\qquad \text{for } x \in [-\pi,\pi] \\
\label{C3C2}
&C_3 = \frac{3}{2} \pi^{1/2} \epsup \delup^{-1} \, , \quad
C_2 = \frac{2 \pi^2 \delup}{3 \lamfac^2} 
\end{align}
The constants $C_3$ and $C_2$ do {\em not} change value crossing the ADC and are given in terms of the dimensionless parameters
\begin{align}
&\epsup = \etastd / \dup
&&\mbox{\em upstream amplitude-to-depth ratio} \\
&\delup = \dup / \lamup
&&\mbox{\em upstream depth-to-wavelength ratio}
\end{align}
The reason for the subscripts $3$ and $2$ will become evident in the next section.

Meanwhile, the dimensionless depth $\drat = {\depth}/{\dup}$ {\em does} change value across the ADC since the depth $\depth$ changes. Recall that the reference frame of \eqref{dimlessKdV} moves with the local wave speed via the variable $\xi = x-ct$ from \eqref{KdV}. Thus, the ADC is met at some time $T_{ADC}$, and for simplicity we set $T_{ADC} = 0$. Therefore, we can regard $\drat$ is a piece-wise constant function of dimensionless time
\begin{equation}
\label{dratpw}
\drat = 
\begin{cases}
1 		&\quad \mbox{for } {t}<0 \\
\dratdn = {\ddn}/{\dup} 	&\quad \mbox{for } {t}>0
\end{cases}
\end{equation}
See Table \ref{paramtable} for a summary of these dimensionless parameters and their values in experiments.

@@

The momentum and energy defined in \eqref{MomEn} remain invariants of TKdV, with the same normalization values $\Mo[\uL] = 0$ and $\En[\uL] = 1$. Note that Parseval's identity implies
\begin{equation}
\En[\uL] = 2 \pi \sum_{k=1}^{\Lambda} \abs{\uhat_k}^2 = 1
\end{equation}
% The above agrees with the first line on top of p. 4 in Majda and Qi J Stat Phys 2019.
Thus, the dynamics of interest are confined to the unit hypersphere, $\En = 1$ in $\CC^{\Lambda}$.
In summary, the TKdV system possesses three important invariants: momentum, energy, and Hamiltonian. The untruncated KdV system possesses an infinite sequence of additional invariants \cite{lax1975periodic, whitham2011linear}, but their truncated counterparts are not generally invariants of TKdV.

@@

	This statistical matching condition, originally derived by Majda et al., imposes a relationship between the two inverse temperatures $\thup$ and $\thdn$. In particular, we view $\thup$ as given by the random-state of the incoming wave field. Thus, while we will treat $\thup$ as a free parameter that can be varied to study various possible system states, the downstream $\thdn$ is determined directly by the matching condition \eqref{statmatch}, giving the functional dependence
\begin{equation}
\thdn = \transf \left( \thup \right)
\qquad \mbox{\em Transfer function}
\end{equation}
The {\em transfer function} $\transf$ will be a key link needed to relate the theory to experiments.

@@

	Since the inverse temperature, $\theta$, is a key modeling parameter, it would be highly desirable to normalize the system such that the value of $\theta$ does not depend sensitively on the truncation number, $\Lambda$, as $\Lambda$ grows large. That way, $\theta$ can be interpreted as a real physical parameter that can be linked to the experiments and whose value does not depend sensitively on where one chooses to truncate the system. The one modeling parameter that is left to be set is $\lamfac$, which represents the number of characteristic wavelengths in the periodic domain. In what follows, we will determine reasonable constraints on $\lamfac$ that allow $\theta$ to be asymptotically independent of $\Lambda$.

@@

	We now aim to compare the wave statistics measured in experiments against those that emerge from the TKdV theory. Throughout, we will focus on the representative set of experiments detailed in Table \ref{paramtable}, in which the peak downstream skewness was measured to be 0.83. 
%We can then use Fig.~\ref{transfig}(c) as a guide to select the appropriate upstream inverse temperature $\thup$. With $(\Lambda, \lamfac) = (16, 8)$ fixed, the figure indicates a value of $\thup = 17$ achieves the desired outgoing skewness. We thus run the deterministic TKdV simulations with model parameters $\Lambda = 16$, $\lamfac = 8$, $\thup = 17$ and with $\epsup$, $\delup$, and $\dratdn$ set to the representative values in Table \ref{paramtable}. 
%
With $(\Lambda, \lamfac) = (16, 8)$ fixed, Fig.~\ref{transfig}(c) indicates $\thup = 20$ as a reasonable value to attain the desired skewness. We thus run the deterministic TKdV simulations these model parameters and with $\epsup$, $\delup$, and $\dratdn$ set to the representative values in Table \ref{paramtable}. We run the simulations to a sufficiently long dimensionless time of $t_{f} = 10$, with $dt = 5 \times 10^{-4}$ and with $10^3$ trajectories sampled from the upstream Gibbs measure $\Gup$.

@@

	Distinct differences in the shapes of the downstream distributions, however, are also visible. While the theoretical distributions remain symmetric, the experimental distributions skew towards negative slope. As mentioned earlier, this skewness is consistent with a right-moving wave of steep leading surface (i.e.~negative slope), which is a phenomenon apparently not captured by the TKdV simulations.
Finally, it should be commented that the experimental measurements of slope involve numerical differentiation of the free surface extracted from optical images, which naturally introduces error. Furthermore, high-order statistics, such as kurtosis, can be very sensitive to noise and measurement error. Hence, the experimental statistics of slope, while highly valuable, should be regarded with some level of caution.

@@

	We remind the reader that the KdV framework tracks the long-time evolution of waves in a reference frame moving with the characteristic speed $c = \sqrt{g \depth}$ from linear theory. Hence, the above discussion of left-or-right going waves cannot be interpreted in the context of the experiments without an appropriate Galilean transformation. Hence, in the laboratory frame, all waves indeed propagate unidirectionally, from left to right, with a speed near $c$. The directions and speeds calculated by the TKdV framework simply quantify the deviation of the true wave speed from $c$.

@@

	These estimates raise an important question: if a distance of 22 meters is required for significant wave evolution under the KdV framework, how do the experiments exhibit substantial changes in wave statistics a much shorter distance downstream of the ADC? The results shown in Fig.~\ref{skew-kurt} help resolve this question. This figure shows the evolution of surface-displacement skewness (left) and kurtosis (right) computed by the same deterministic TKdV simulations as pictured in Fig.~\ref{trajectories}. The bold black curve shows the ensemble mean (ensemble size of 1000), while the faint gray curves show the skewness and kurtosis of 100 individual trajectories to give a sense for the variation involved. Importantly, the skewness and kurtosis evolve on a much shorter timescale than the aforementioned $t=7$. More precisely, skewness and kurtosis have already saturated to their asymptotic value by $t = 1$, and reach half of that value by $t = 0.18$. These dimensionless times correspond to travel distances of 310 cm and 56 cm respectively --- much shorter than the previously mentioned 22 meters, and on the same order as the relevant distances in the experiments. Thus, after crossing the ADC, the wave-field rapidly reconfigures itself enough to fundamentally alter its statistical distributions. The timescale for this reconfiguration is much shorter than the time required for waves to cross the entire periodic domain.

@@

	This manuscript extends the parallel experimental and modeling efforts of Bolles et al.~\cite{bolles2019} and Majda et al.~\cite{majda2019} concerning the emergence of anomalous wave statistics from abrupt changes in bottom topography. The theoretical framework is based on deterministic and statistical analysis of the TKdV system, with exploitation of the Hamiltonian structure and the associated invariant Gibbs measures. The theory depends crucially on matching the incoming and outgoing invariant measures at the abrupt depth change, so as to link the outgoing dynamics to the incoming ones. Throughout, we have emphasized the synergy between the experiments and theory, with experimental data informing theoretical advancements and model predictions motivating new experimental measurements.

>>>>

\title{Anomalous waves triggered by abrupt depth changes: laboratory experiments and truncated KdV statistical mechanics}

@@

A recent series of laboratory \cite{bolles2019} and numerical investigations \cite{viotti2014} have demonstrated the emergence of anomalous wave statistics from abrupt variations in bottom topography. Since topographical variations are strictly one-dimensional, these studies can be viewed as offering a bare minimum set of conditions capable of generating anomalous waves. In particular, the more complex mechanism of focusing by 2D bottom topography absent. The studies thus offer a paradigm system, with emergent anomalous statistical properties similar to that seen in more complex systems, but in a tractable context that is amenable to analysis.

@@

We consider the surface displacement $\eta(x,t)$ of unidirectional, shallow-water waves in a reference frame moving with the characteristic wave speed, $\xi = x - ct$. Here, $c = \sqrt{g \depth}$ is leading-order approximation to the wave speed (i.e.~from linear theory), where $g$ is gravity and $\depth$ is the local depth.
The leading-order dynamics are corrected to first order in small amplitude by the Korteweg–de Vries equation (KdV), which in dimensional form is given by \cite{whitham2011linear}
\begin{equation}
\label{KdV}
\eta_t + \frac{3 c}{2 \depth} \eta \eta_{\xi} + \frac{c \depth^2}{6} \eta_{\xi \xi \xi} = 0
\end{equation}
% Given by Whitham on p. 461, Eq. 13.98
Motivated by the experiments, we consider waves that originate from a region of constant depth, encounter an abrupt depth change, and continue into another region of constant depth. Thus, depth will be piecewise constant
\begin{align}
\depth = 
\begin{cases}
\dup \quad \mbox{if } x<0 \\
\ddn \quad \mbox{if } x>0
\end{cases}
\end{align}
Most often, we will consider waves moving into shallower depth, so that $\dup > \ddn$.

@@

With the above choices, the dimensionless KdV equation takes the form
\begin{align}
\label{dimlessKdV}
&u_t + C_3 \drat^{-3/2} \, u u_x + C_2 \drat^{1/2} \, u_{xxx} = 0
\qquad \text{for } x \in [-\pi,\pi] \\
\label{C3C2}
&C_3 = \frac{3}{2} \pi^{1/2} \epsup \delup^{-1} \, , \quad
C_2 = \frac{2 \pi^2 \delup}{3 \lamfac^2} 
\end{align}
The constants $C_3$ and $C_2$ do {\em not} change value crossing the ADC and are given in terms of the dimensionless parameters
\begin{align}
&\epsup = \etastd / \dup
&&\mbox{\em upstream amplitude-to-depth ratio} \\
&\delup = \dup / \lamup
&&\mbox{\em upstream depth-to-wavelength ratio}
\end{align}
The reason for the subscripts $3$ and $2$ will become evident in the next section.

Meanwhile, the dimensionless depth $\drat = {\depth}/{\dup}$ {\em does} change value across the ADC since the depth $\depth$ changes. Recall that the reference frame of \eqref{dimlessKdV} moves with the local wave speed via the variable $\xi = x-ct$ from \eqref{KdV}. Thus, the ADC is met at some time $T_{ADC}$, and for simplicity we set $T_{ADC} = 0$. Therefore, we can regard $\drat$ is a piece-wise-constant function of dimensionless time
\begin{equation}
\label{dratpw}
\drat = 
\begin{cases}
1 		&\quad \mbox{for } {t}<0 \\
\dratdn = {\ddn}/{\dup} 	&\quad \mbox{for } {t}>0
\end{cases}
\end{equation}
See Table \ref{paramtable} for a summary of these dimensionless parameters and their values in experiments.

@@

The momentum and energy defined in \eqref{MomEn} remain invariants of TKdV, with the same normalized values $\Mo[\uL] = 0$ and $\En[\uL] = 1$. Note that Parseval's identity implies
\begin{equation}
\En[\uL] = 2 \pi \sum_{k=1}^{\Lambda} \abs{\uhat_k}^2 = 1
\end{equation}
% The above agrees with the first line on top of p. 4 in Majda and Qi J Stat Phys 2019.
Thus, the dynamics of interest are confined to the unit hypersphere, $\En = 1$ in $\CC^{\Lambda}$.
In summary, the TKdV system possesses three important invariants: momentum, energy, and Hamiltonian. The untruncated KdV equation possesses an infinite sequence of additional invariants \cite{lax1975periodic, whitham2011linear}, but their truncated counterparts are not generally invariants of TKdV.

@@

	This statistical matching condition, originally derived by Majda et al.~(2019) \cite{majda2019}, imposes a relationship between the two inverse temperatures $\thup$ and $\thdn$. In particular, we view $\thup$ as given by the random-state of the incoming wave field. Thus, while we will treat $\thup$ as a free parameter that can be varied to study various possible system states, the downstream $\thdn$ is determined directly by the matching condition \eqref{statmatch}, giving the functional dependence
\begin{equation}
\thdn = \transf \left( \thup \right)
\qquad \mbox{\em Transfer function}
\end{equation}
The {\em transfer function} $\transf$ will be a key link needed to relate the theory to experiments.

@@

	Since the inverse temperature, $\theta$, is a key modeling parameter, it would be highly desirable to normalize the system such that the value of $\theta$ does not depend sensitively on the truncation index, as $\Lambda$ grows large. That way, $\theta$ can be interpreted as a real physical parameter that can be linked to the experiments and whose value does not depend sensitively on where one chooses to truncate the system. The one modeling parameter that is left to be set is $\lamfac$, which represents the number of characteristic wavelengths in the periodic domain. In what follows, we will determine reasonable constraints on $\lamfac$ that allow $\theta$ to be asymptotically independent of $\Lambda$.

@@

	We now aim to compare the wave statistics measured in experiments against those that emerge from the TKdV theory. Throughout, we will focus on the representative set of experiments detailed in Table \ref{paramtable}, in which the peak downstream skewness was measured to be 0.83. 
%We can then use Fig.~\ref{transfig}(c) as a guide to select the appropriate upstream inverse temperature $\thup$. With $(\Lambda, \lamfac) = (16, 8)$ fixed, the figure indicates a value of $\thup = 17$ achieves the desired outgoing skewness. We thus run the deterministic TKdV simulations with model parameters $\Lambda = 16$, $\lamfac = 8$, $\thup = 17$ and with $\epsup$, $\delup$, and $\dratdn$ set to the representative values in Table \ref{paramtable}. 
%
With $(\Lambda, \lamfac) = (16, 8)$ fixed, Fig.~\ref{transfig}(c) indicates $\thup = 20$ as a reasonable value to attain the desired skewness. We thus run the deterministic TKdV simulations with these model parameters and with $\epsup$, $\delup$, and $\dratdn$ set to the representative values in Table \ref{paramtable}. We run the simulations to a sufficiently long dimensionless time of $t_{f} = 10$, with $dt = 5 \times 10^{-4}$ and with $10^3$ trajectories sampled from the upstream Gibbs measure $\Gup$.

@@

	Distinct differences in the shapes of these downstream distributions, however, are also visible. While the theoretical distributions remain symmetric, the experimental distributions skew towards negative slope. As mentioned earlier, this skewness is consistent with a right-moving wave of steep leading surface (i.e.~negative slope), which is a phenomenon apparently not captured by the TKdV simulations.
Finally, it should be commented that the experimental measurements of slope involve numerical differentiation of the free surface extracted from optical images, which naturally introduces error. Furthermore, high-order statistics, such as kurtosis, can be very sensitive to noise and measurement error. Hence, the experimental statistics of slope, while highly valuable, should be regarded with some level of caution.

@@

	We remind the reader that the KdV framework tracks the long-time evolution of waves in a reference frame moving with the characteristic speed $c = \sqrt{g \depth}$ from linear theory. Hence, the above discussion of left-or-right going waves cannot be interpreted in the context of the experiments without an appropriate Galilean transformation. Thus, in the laboratory frame, all waves indeed propagate unidirectionally, from left to right, with a speed near $c$. The directions and speeds calculated by the TKdV framework simply quantify the deviation of the true wave speed from $c$.

@@

	These estimates raise an important question: if a distance of 22 meters is required for significant wave evolution under the KdV framework, how do the experiments exhibit substantial changes in wave statistics a much shorter distance downstream of the ADC? The results shown in Fig.~\ref{skew-kurt} help resolve this question. This figure shows the evolution of surface-displacement skewness (left) and kurtosis (right) computed by the same deterministic TKdV simulations as pictured in Fig.~\ref{trajectories}. The bold black curve shows the ensemble mean (ensemble size of 1000), while the faint gray curves show the skewness and kurtosis of 100 individual trajectories to give a sense for the variation involved. Importantly, the skewness and kurtosis evolve on a much shorter timescale than the aforementioned $t=7$. More precisely, skewness and kurtosis have already saturated to their asymptotic value by $t = 1$, and reach half of that value by $t = 0.18$. These dimensionless times correspond to travel distances of 310 cm and 56 cm respectively --- much shorter than the previously mentioned 22 meters, and on the same order as the relevant distances in the experiments. Thus, after crossing the ADC, the wave-field rapidly reconfigures itself enough to fundamentally alter its statistical distributions, and the timescale for this reconfiguration is much shorter than the time required for waves to cross the entire periodic domain.

@@

	This manuscript extends the parallel experimental and modeling efforts of Bolles et al.~\cite{bolles2019} and Majda et al.~\cite{majda2019} concerning the emergence of anomalous wave statistics from abrupt changes in bottom topography. The theoretical framework is based on deterministic and statistical analysis of the TKdV system, with exploitation of the Hamiltonian structure and the associated invariant Gibbs measures. The theory depends crucially on matching the incoming and outgoing invariant measures at the abrupt depth change, so as to link the incoming and outgoing states. Throughout, we have emphasized the synergy between the experiments and theory, with experimental data informing theoretical advancements and model predictions motivating new experimental measurements.
